### 2021-09-10 documents/sections/proto_carib.tex
added split-s illustration data, more examples

\begin{table}
	\centering
	\caption{Participles of \gl{s_a_} and \gl{s_p_} verbs}
	\label{tab:participles}
	\begin{tabular}{@{}llll@{}}
	\mytoprule
Language & \gl{s_a_} verb & \gl{s_p_} verb & Source(s)\\
	\mymidrule
\kaxui & \obj{t-ehurka-t͡ʃe} \qu{fallen} & \obj{tɨ-jaʔ-so} \qu{burnt} & \textcite[39]{schuring2018kaxuyana} \\
\arara & \obj{t-\emp{o-}ep-te} \qu{come} & \obj{t-oregrum-te} \qu{sad} & \textcite[207, 118]{alves2017arara} \\
\trio & \obj{tɨ-\emp{w-}əturu-e} \qu{talked} & \obj{t-əpəə-se} \qu{tired} & \textcite[333--334]{triomeira1999} \\
\wayana & \obj{tə-\emp{w-}epɨ-he} \qu{bathed} & \obj{t-onopɨ-he} \qu{painted} & \textcite[400]{wayanatavares2005} \\
\kalina & \obj{tu-\emp{w-}oʔka-se} \qu{come out} & \obj{t-okari-se} \qu{told} & \textcite[433, 430]{hoff1968carib} \\
\apalai & \obj{t-\emp{o-}erama-se} \qu{returned} & \obj{t-anorɨ-se} \qu{dried} & \textcite[42, 30]{camargo2002lexico}\\
\panare & \obj{t-\emp{o-}tatɨhpə-se} \qu{wailed} & \obj{tɨ-sɨrɨke-t͡ʃe} \qu{tired} & \textcite[232, 244]{panarepayne2013} \\
	\mybottomrule
	\end{tabular}
\end{table}

@@

\subsection{Person marker extensions in intransitive verbs}
\label{sec:extensions_intro}
The original \PC split-\gl{s} system has been subject to change in many languages, in some even to the point of total loss.
Besides such modifications as for example losing the \gl{s_a_} class marker \rc{w-} \pcref{sec:split} or losing certain person values altogether, these changes have largely been due to person prefixes being extended to new verbs.\footnote{
One could also imagine a scenario whereby verbs of one of the classes are gradually replaced with innovative verbs from the other class, as suggested for \panare by \textcite[225]{meira2000split}.
However, the verbs he sampled came from the \emp{A}-section of \posscite{mattei1994diccionario} dictionary \perscommpar{Spike Gildea}, and \obj{at-}\slash{}\obj{at͡ʃ-}\slash{}\obj{as-} is a frequent reflex of \detrz in \panare, meaning that many \gl{s_a_} verbs are \obj{a}-initial and explaining the 80:40 \gl{s_a_}:\gl{s_p_} ratio in Meira's sample.
Thus, the hypothesis that \panare has largely replaced \gl{s_p_} with \gl{s_a_} verbs remains to be thoroughly tested.
For the moment, the most likely scenario leading to the eventual loss of the split-\gl{s} system includes markers being extended to new verbs.}
There have been many such person marker extensions in Cariban languages, and some are still ongoing.
This is shown by \textcite{gildea1998}, using the Parukotoan languages as an example.
I have reproduced his tables as a tree diagram in \cref{fig:par_ext}, with adapted transcription -- and in the case of \kaxui, the addition of Ø/\obj{j-} as an alternative \gl{1}\gl{s_p_} marker \perscommpar{Spike Gildea}.
Apart from segmental changes to individual morphemes, the following restructuring innovations happened in the \setone paradigm in Parukotoan:

@@

\begin{table}[h]
\caption{Some examples for completed extensions \parencite[90--92]{gildea1998}}
\label{tab:completed}
\begin{subtable}[h]{0.24\textwidth}
\centering
\caption{\apalai}
\label{tab:apalai}
\begin{tabular}{@{}lll@{}}
\mytoprule
& \gl{s_a_} & \gl{s_p_}\\
\mymidrule
\gl{1} & Ø/\obj{ɨ-} & \obj{ɨ-}/\obj{j-}\\
\gl{2} & \obj{m(ɨ)-} & \obj{o-}\\
\gl{1+2} & \multicolumn{2}{c}{\emp{\obj{s(ɨ)-}}}\\
\gl{3} & \multicolumn{2}{c}{\obj{n(ɨ)-}}\\
\mybottomrule
\end{tabular}
\end{subtable}
\hfill
\begin{subtable}[h]{0.24\textwidth}
\centering
\caption{\panare}
\label{tab:panare}
\begin{tabular}{@{}lll@{}}
\mytoprule
& \gl{s_a_} & \gl{s_p_}\\
\mymidrule
\gl{1} & \obj{w(ɨ)-} & Ø/\obj{j-}\\
\gl{2} & \multicolumn{2}{c}{\emp{\obj{m(ɨ)-}}}\\
\gl{1+2} & \multicolumn{2}{c}{\obj{n(ɨ)-}}\\
\gl{3} & \multicolumn{2}{c}{\obj{n(ɨ)-}}\\
\mybottomrule
\end{tabular}
\end{subtable}
\hfill
%\begin{subtable}[h]{0.24\textwidth}
%\centering
%\caption{\pemon}
%\label{tab:pemon}
%\begin{tabular}{@{}lll@{}}
%\mytoprule
%& \gl{s_a_} & \gl{s_p_}\\
%\mymidrule
%\gl{1} & \multicolumn{2}{c}{\emp{Ø}}\\
%\gl{2} & \multicolumn{2}{c}{\emp{\obj{m-}}}\\
%\gl{1+2} & \multicolumn{2}{c}{\obj{n-}}\\
%\gl{3} & \multicolumn{2}{c}{\obj{n-}}\\
%\mybottomrule
%\end{tabular}
%\end{subtable}
%\hfill
\begin{subtable}[h]{0.24\textwidth}
\centering
\caption{\waimiri}
\label{tab:waimiri}
\begin{tabular}{@{}ll@{}}
\mytoprule
& \gl{s}\\
\mymidrule
\gl{1} & \emp{\obj{w(ɨ)-}/\obj{i-}}\\
\gl{2} & \emp{\obj{m(ɨ)-}}\\
\gl{1+2} & \emp{\obj{h(ɨ)-}}\\
\gl{3} & \obj{n-}/Ø\\
\mybottomrule
\end{tabular}
\end{subtable}
\end{table}

>>>>

\begin{table}
	\centering
	\caption{Participles of \gl{s_a_} and \gl{s_p_} verbs \parencites[118, 207]{alves2017arara}[30, 42]{camargo2002lexico}[430, 433]{hoff1968carib}[333, 334]{triomeira1999}[400]{wayanatavares2005}[39]{schuring2018kaxuyana}[232, 244]{panarepayne2013}}
	\label{tab:participles}
\input{floats/participles}
\end{table}

\begin{table}
	\centering
	\caption{Nominalizations of \gl{s_a_} and \gl{s_p_} verbs \parencites[49, 74]{schuring2018kaxuyana}[97]{alves2017arara}[246]{triomeira1999}[130, 409]{wayanatavares2005}[90]{koehn1986apalai}[ner2-003]{koehns1994textos}[135, 392]{hoff1968carib}[390]{panarepayne2013}[23]{mattei1994diccionario}}
	\label{tab:nmlz}
	\input{floats/nominalizations}
\end{table}

@@

\subsection{Person marker extensions in intransitive verbs}
\label{sec:extensions_intro}
The original \PC split-\gl{s} system has been subject to change in many languages, in some even to the point of total loss.
Besides such modifications as for example losing the \gl{s_a_} class marker \rc{w-} \pcref{sec:split} or losing certain person values altogether, these changes have largely been due to person prefixes being extended to new verbs.\footnote{
One could also imagine a scenario whereby verbs of one of the classes are gradually replaced with innovative verbs from the other class, as suggested for \panare by \textcite[225]{meira2000split}.
However, the verbs he sampled came from the \emp{A}-section of \posscite{mattei1994diccionario} dictionary \perscommpar{Spike Gildea}, and \obj{at-}\slash{}\obj{at͡ʃ-}\slash{}\obj{as-} is a frequent reflex of \detrz in \panare, meaning that many \gl{s_a_} verbs are \obj{a}-initial and explaining the 80:40 \gl{s_a_}:\gl{s_p_} ratio in Meira's sample.
Thus, the hypothesis that \panare has largely replaced \gl{s_p_} with \gl{s_a_} verbs remains to be thoroughly tested.
For the moment, the most likely scenario leading to the eventual loss of the split-\gl{s} system includes markers being extended to new verbs.}
There have been many such person marker extensions in Cariban languages, and some are still ongoing.
This is shown by \textcite{gildea1998}, using the Parukotoan languages as an example.
I have reproduced his tables as a tree diagram in \cref{fig:par_ext}, with adapted transcription -- and in the case of \kaxui, the addition of ∅/\obj{j-} as an alternative \gl{1}\gl{s_p_} marker \perscommpar{Spike Gildea}.
Apart from segmental changes to individual morphemes, the following restructuring innovations happened in the \setone paradigm in Parukotoan:

@@

\begin{table}[h]
\caption{Some examples for completed extensions \parencite[90--92]{gildea1998}}
\label{tab:completed}
\begin{subtable}[h]{0.24\textwidth}
\centering
\caption{\apalai}
\label{tab:apalai}
\begin{tabular}{@{}lll@{}}
\mytoprule
& \gl{s_a_} & \gl{s_p_}\\
\mymidrule
\gl{1} & ∅/\obj{ɨ-} & \obj{ɨ-}/\obj{j-}\\
\gl{2} & \obj{m(ɨ)-} & \obj{o-}\\
\gl{1+2} & \multicolumn{2}{c}{\emp{\obj{s(ɨ)-}}}\\
\gl{3} & \multicolumn{2}{c}{\obj{n(ɨ)-}}\\
\mybottomrule
\end{tabular}
\end{subtable}
\hfill
\begin{subtable}[h]{0.24\textwidth}
\centering
\caption{\panare}
\label{tab:panare}
\begin{tabular}{@{}lll@{}}
\mytoprule
& \gl{s_a_} & \gl{s_p_}\\
\mymidrule
\gl{1} & \obj{w(ɨ)-} & ∅/\obj{j-}\\
\gl{2} & \multicolumn{2}{c}{\emp{\obj{m(ɨ)-}}}\\
\gl{1+2} & \multicolumn{2}{c}{\obj{n(ɨ)-}}\\
\gl{3} & \multicolumn{2}{c}{\obj{n(ɨ)-}}\\
\mybottomrule
\end{tabular}
\end{subtable}
\hfill
%\begin{subtable}[h]{0.24\textwidth}
%\centering
%\caption{\pemon}
%\label{tab:pemon}
%\begin{tabular}{@{}lll@{}}
%\mytoprule
%& \gl{s_a_} & \gl{s_p_}\\
%\mymidrule
%\gl{1} & \multicolumn{2}{c}{\emp{∅}}\\
%\gl{2} & \multicolumn{2}{c}{\emp{\obj{m-}}}\\
%\gl{1+2} & \multicolumn{2}{c}{\obj{n-}}\\
%\gl{3} & \multicolumn{2}{c}{\obj{n-}}\\
%\mybottomrule
%\end{tabular}
%\end{subtable}
%\hfill
\begin{subtable}[h]{0.24\textwidth}
\centering
\caption{\waimiri}
\label{tab:waimiri}
\begin{tabular}{@{}ll@{}}
\mytoprule
& \gl{s}\\
\mymidrule
\gl{1} & \emp{\obj{w(ɨ)-}/\obj{i-}}\\
\gl{2} & \emp{\obj{m(ɨ)-}}\\
\gl{1+2} & \emp{\obj{h(ɨ)-}}\\
\gl{3} & \obj{n-}/∅\\
\mybottomrule
\end{tabular}
\end{subtable}
\end{table}
